Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-29 manuscript/manuscript.tex
Edit the discussion
Rename to just Discussion. Remove some of the things repeated in the
intro about Lin2018 and Fukushima2018. I kept the part that mentions
that we could use the density discretization combined with piecewise
linear or polynomial functions (citing both papers).
Added some parts about why our method is independent of the density
function. Will reiterate this in the conclusions as well.
Argue why we can probably extrapolate the results here to any realistic
density function.

We computed the differences between the results for the exponential density function and
those generated by the same model but now with a constant density (the mean of
-412kg/m$^3$ and -275kg/m$^3$).
We also computed the differences between the exponential density and a linear density
that assumes these two values on the highest and the deepest point of the basin
(Fig.~\ref{fig:neuquen-basin-densities}).
Fig.~\ref{fig:neuquen-basin-diffs}a-b and Fig.~\ref{fig:neuquen-basin-diffs}c-d show the
differences with the constant and the linear density, respectively.
The maximum absolute difference in the computed $g_z$ is approximately 8 mGal for the
homogeneous case and approximately 6 mGal for the linear density case.
Both are well above the accuracy of data derived from spherical harmonic models, which
is often used for regional interpretations.

@@

\section{Results and Discussion}

% Brought this down from the results. It fits better in the discussion. I'll find a
% place for it afterwards.
These results also show that there's no meaningful dependency between the accuracy of
the method and the thickness of the shell or the height of the computation grid (the
results for the Satellite grid and the Global grid are very similar).

The $D$ and $\delta$ ratios determinations proved that for a linear density case only
the two dimensional adaptive discretization algorithm and a second-order GLQ
approximation is needed to achieve a 0.1\% accuracy of the gravitational potential and
the vertical component of the gradient ($g_z$) by using a $D=$ 1 and 2.5 for each field,
respectively.
Although for non-linear functions the adaptive discretization is not sufficient and the
density-based discretization algorithm must be applied.
The $D$-$\delta$ space exploration showed that the more efficient computation takes
place when the value of $D$ equals to the ones determined in the linear density
scenario, while the value of $\delta$ must be determined for each specific density
variation.
In case of an exponential density, a $\delta$ equal to 0.1 is sufficient to achieve the
0.1\% accuracy threshold, while a $\delta = 0.01$ is needed for a sinusoidal density
function.

This result could be extrapolated: a default value of $\delta = 0.01$ would guarantee an
accuracy of 0.1\% in case of an arbitrary continuous density function.
Nevertheless, we encourage to perform similar accuracy tests in order to proof that this
extrapolation works for the desired density function, specially for more complex
variations than a polynomial or exponential one.

\citet{Lin2018} proposed a combined method to efficiently compute gravitational fields
generated by tesseroids: it uses a GLQ with a two dimensional adaptive discretization
algorithm for near computation points and a Taylor series expansion method for far
computation points.
Because our goal is to build a method that could be applied to any continuous density
function, the Taylor series wouldn't be well suited for this task and therefore we
focused only on a pure GLQ approximation method.
Nevertheless, our method could be easily combined with a Taylor series expansion for
specific density function, i.e.~a linear variation.

Moreover, \citet{Lin2018} show that the combined method can be applied to a linear
density variation by splitting the gravitational field integral in two: one for each
term of the linear function.
Our results show that in case of a linear density variation a single GLQ computation is
sufficient to accurately compute the fields, thus reducing the computation time by half.

\citet{Fukushima2018} propose a generalization of his method to polynomial density
functions.
On the other hand, our method is not bounded to a specific density function.
Nevertheless, the density-based discretization algorithm could be used before applying
the numerical integration developed by \citet{Fukushima2018} assigning a low polynomial
order density to each discretizing tesseroid.
Although this method would involve a computational error due to the approximation of the
density function as a polynomial one, its feasibly if we account for this error and
a method to limit it.

We have also carried out $D$ and $\delta$ determinations for the vertical component of
the Marussi tensor ($g_{zz}$) for linear and exponential densities.
We didn't include these results because our methodology wouldn't achieve the 0.1\%
accuracy threshold for the general case of such density functions.
Moreover, the values of $D$ and $\delta$ needed for the cases at which the accuracy was
below 0.1\% implied a high number of discretizations, thus greatly increasing the
computation time.
In summary, we found the computation of the Marussi tensor components in case of
variable density tesseroids a subject for future research.

@@

>>>>

We computed the differences between the results for the exponential density function and
those generated by the same model but now with a constant density (the mean of
-412kg/m$^3$ and -275kg/m$^3$).
We also computed the differences between the exponential density and a linear density
that assumes these two values on the highest and the deepest point of the basin
(Fig.~\ref{fig:neuquen-basin-densities}).
Fig.~\ref{fig:neuquen-basin-diffs}a-b and Fig.~\ref{fig:neuquen-basin-diffs}c-d show the
differences with the constant and the linear density, respectively.
The maximum absolute difference in the computed $g_z$ is approximately 8 mGal for the
homogeneous case and approximately 6 mGal for the linear density case.
Both are well above the accuracy of most available data products.

@@

\section{Discussion}

By including the density function into the GLQ integration, the method described here
can be applied without modification to tesseroids with density following any continuous
function of radius.
A density-based discretization algorithm divides the tesseroid in the radial dimension
to ensure accurate integration of the density function.
This algorithm is independent of the GLQ integration and could potentially be used to
determine an optimal discretization when approximating a density function by piecewise
linear \citep{Lin2018} or piecewise polynomial \citep{Fukushima2018} functions.

Our numerical experiments show that the two dimensional adaptive discretization is
enough to achieve 0.1\% accuracy with a second-order GLQ in the case of a linear density
function (Fig.~\ref{fig:D-linear}).
Values of the distance-size ratio determined here for the gravitational potential
($D=1$) and its vertical derivative ($D=2.5$) are compatible with \citet{Uieda2016}.
These results also show that there is no significant relationship between the accuracy
of the method and the thickness of the tesseroid model.

The $D$-$\delta$ space exploration for the exponential density function
(Fig.~\ref{fig:grid-search}) showed that values of $D$ determined for the linear density
case are equally applicable to the exponential density case.
Furthermore, the difference with respect to the analytical solution only falls below
0.1\% for values of $\delta$ below 0.1.
It follows that the density-based discretization is required to achieve the desired
accuracy level for non-linear density functions.

More detailed error analyses showed that $\delta = 0.1$ is sufficient to guarantee an
accuracy of 0.1\% for all exponential density functions tested
(Fig.~\ref{fig:delta-exponential})
and most of the sinusoidal functions tested (Fig.~\ref{fig:delta-sine}).
The exception is the case of a sinusoidal function with $b = 10$
(Eq.~\ref{eq:density-sine}), for which $\delta = 0.01$ is required to achieve 0.1\%
accuracy.
Nevertheless, using $\delta = 0.1$ in this case would still produce results with an
error of less than 1\%.

While the algorithms proposed here can be applied to any continuous density function,
the optimal values for $D$ and $\delta$ are empirically determined only for linear,
exponential, and sinusoidal density functions.
Thus, we can only say with certainty that these values will produce results with 0.1\%
accuracy for these functions.
However, all of our numerical tests include worse-case scenarios (zero computation
heights, large tesseroids, highly variable density functions, etc).
For this reason, these optimal values of $D$ and $\delta$ can plausibly be
extrapolated to any realistic continuous density function.
Notwithstanding, we encourage users of the algorithms and software to perform similar
tests in order evaluate the accuracy when using density functions more complex than the
ones tested here.

@@

% Moved here from the discussion. The conclusions are the place to talk about future
% work and speculations.
We have also carried out $D$ and $\delta$ determinations for the vertical component of
the Marussi tensor ($g_{zz}$) for linear and exponential densities.
We didn't include these results because our methodology wouldn't achieve the 0.1\%
accuracy threshold for the general case of such density functions.
Moreover, the values of $D$ and $\delta$ needed for the cases at which the accuracy was
below 0.1\% implied a high number of discretizations, thus greatly increasing the
computation time.
In summary, we found the computation of the Marussi tensor components in case of
variable density tesseroids a subject for future research.
